Effect of initial volume fraction on the collapse of granular columns in fluid

### 2018-08-27 paper/paper.tex
:art: Fix geometry setup without slope
:computer: Draft paper for IS-Atlanta2

The flow of dense granular material is a common phenomenon in engineering predictions, such as avalanches, landslides, and debris-flow modelling. Despite the huge amount of research that has gone into describing the behaviour of granular flows, a constitutive equation that describes the overall behaviour of a flowing granular material is still lacking. The initiation and propagation of submarine granular flows depend mainly on the slope, density, and quantity of the material destabilised. Although certain macroscopic models are able to capture the simple mechanical behaviours, the complex physical mechanisms that occur at the grain scale, such as thydrodynamic instabilities, the formation of clusters, collapse, and transport, have largely been ignored~\shortcite{Topin2011}. The momentum transfer between the discrete and the continuous phases significantly affects the dynamics of the flow~\shortcite{Peker2007}. Grain-scale description of the granular material enriches the macro-scale variables,  which poorly account for the local rheology of the materials.  In order to describe the mechanism of saturated and/or immersed granular flows, it is important to consider both the dynamics of the solid phase and the role of the ambient fluid~\shortcite{Denlinger2001}. In particular, when the solid phase reaches a high volume fraction, it is important to consider the strong heterogeneity arising from the contact forces between the grains, the drag interactions which counteract the movement of the grains, and the hydrodynamic forces that reduce the weight of the solids inducing a transition from dense compacted to a dense suspended flow~\shortcite{Meruane2010}. The case of the collapse in presence of an interstitial fluid has been less studied. In this paper, we study the submarine granular flows in the inclined configuration. We study the effect of permeability, density and slope angle on the run-out evolution.

@@

The Lattice Boltzmann Method is a `micro-particle' based numerical time-stepping procedure for the solution of incompressible fluid flows. Consider a 2D incompressible fluid flow with density $\rho$ and kinematic viscosity \textit{v}, in a rectangular domain \textit{\textbf{D}}. The fluid domain is divided into a rectangular grid or lattice, with the same spacing \textit{`h'} in both the \textit{x-} and the \textit{y-}directions~\shortcite{Kumar2015}. The present study focuses on two-dimensional problems, hence the \textit{D2Q9} momentum discretisation is adopted (see \shortciteN{He1997} for naming convention). The Multiple Relaxation Time LBM (LBM-MRT) is adopted in the present study for numerical stability. Further details about the modelling can be found in~\shortciteNP{Kumar2015,Kumar2017}.

@@

\section{UNDERWATER GRANULAR FLOWS}
In this study, a 2D poly-disperse system ($d_{max}/d_{min} = 1.8$) of circular discs in fluid was used to understand the behaviour of granular flows on inclined planes (see~\Cref{fig:setup}). The soil column was modelled using 1000 discs of density \SI{2650}{\kg\per\cubic\meter} and a contact friction angle of \SI{26}{\degree}. The collapse of the column was simulated inside a fluid with a density of \SI{1000}{\kg\per\cubic\meter}  and a kinematic viscosity of \SI{1e-6}{\square\meter\per\second}. The choice of a 2D geometry has the advantage of cheaper computational effort than a 3D case, making it feasible to simulate very large systems. A granular column of aspect ratio `a' of 0.8 was used. Dry analyses were also performed to study the effect of hydrodynamic forces on the run-out distance.

@@

\shortciteNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), is compared with a loose
granular column (79\% packing fraction). The permeability is varied by changing
the hydrodynamic radius from 0.7 \textit{R} to 0.95 \textit{R}.

@@

At a high permeability (\textit{r} = 0.7 \textit{R}), the evolution of run-out 
at the initial stage is quicker, which means that the negative pore-pressure 
that is developed during the shearing along the shear-failure surface is 
dissipated faster. Even though the negative pore-pressure is dissipated, due to 
the development of negative pore-pressure the evolution of run-out in fluid is 
slower than its dry counterpart. The rate of pore-pressure dissipation 
decreases with decrease in the permeability. This can be observed by a flatter 
slope in the run-out evolution with decrease in the 
permeability.~\Cref{fig:PWP_ini_dense} shows the distribution of pore-pressure 
in high and low permeability granular media along the horizontal direction at a 
height of $10 \times d$ from the base. In LBM, the pore-pressure in fluid is a 
function of fluid density distribution functions. At time $t = \tau_c$, the 
highly permeable (\textit{r} = 0.7 \textit{R}) granular column 
shows smaller negative pore-pressure in comparison to large negative 
pore-pressures observed in the shearing zone of a low permeable column 
(\textit{r} = 0.9 R). This shows that not only does it take longer for the 
pore-pressure to dissipate with a decrease in permeability, but also results in 
almost twice the negative pore-pressure than what is observed in the high 
permeable case (\cref{fig:r095_PWP_ini_dense}). A 
high value of positive pore-pressure is observed at the face of the column 
in the low permeable condition. This indicates that the low permeable column 
fails as a continuous block undergoing a shear failure, which generates a very 
large negative pore-pressure along the failure surface. However, in the case of 
the high permeable column the failure is more localised with multiple negative 
pore-pressure spikes (\cref{fig:r07_PWP_ini_dense}).

@@

The high permeability granular column shows lower water entrainment 
(\cref{fig:Packing_Density_a08_dense}), which indicates that for highly 
permeable flows the drag force acting on the soil grains predominates over the 
lubrication effect on the run-out behaviour.~\Cref{fig:Fr_a08_dense} shows the 
evolution of Froude's number with time for different permeability conditions. 
In dense granular columns, the Froude's number increases with decrease in 
permeability, however the Froude's number is below a critical value of 0.4 even 
for a low permeable column. Hence, no hydroplaning is observed. In both the low 
and high permeable granular flows, the granular material consolidates at the 
final stage of the flow (\cref{fig:Packing_Density_a08_dense}). This can be 
observed by the increase in the packing density at the final stage due to 
settlement of grains and expulsion of entrained water.

@@

The normalised run-out evolution with time for a loose initial packing (79\% 
packing fraction) with different hydrodynamic radii 0.7 \textit{R}, 0.8 
\textit{R}, 0.9 and 0.95 R are presented in~\cref{fig:Runout_a08_loose}. The 
run-out evolution a column of grains in suspension is compared with the dry and 
submerged granular columns to understand the influence of hydrodynamic forces 
on the flow kinematics. Similar to the dense granular column, the run-out 
distance increases with increase in the hydrodynamic radius (i.e., decrease in 
permeability). At low permeabilities (\textit{r} = 0.9 and 0.95 \textit{R}), 
the run-out distance is longer than the dry condition. This shows that the 
lubrication effect in low permeability conditions overcomes the influence of 
the drag force and the development of large negative pore-pressure resulting in 
a longer run-out distance. Although the suspended granular masses experience 
higher drag forces and turbulent effects, the run-out evolves almost at the 
same rate in comparison with granular columns with high permeability. This 
shows the effect of permeability on the dissipation rate of negative 
pore-pressure developed during the initial stage of collapse.

@@

Two-dimensional LB-DEM simulations were performed to understand the behaviour of
submarine granular flows. Unlike dry granular collapse, the run-out behaviour in
fluid is dictated by the initial volume fraction.
\shortciteNP{Rondon2011} also observed that the collapse of a granular column in a 
viscous fluid is mainly controlled by the initial volume fraction and not by 
the aspect ratio of the column. The role of the initial volume fraction 
observed explains the pore pressure feedback mechanism proposed 
by~\shortciteANP{Schaeffer2008} in the context of landslides. The dense column needs to 
dilate in order to flow. When it starts to fall, liquid is then sucked into the 
column, which is then stabilised by the additional viscous 
drag~\shortcite{Rondon2011,Topin2012}. By opposition the loose column when 
it starts flowing expands and ejects liquid, leading to a partial fluidisation 
of the material.

>>>>

The flow of dense granular material is a common phenomenon in engineering predictions, such as avalanches, landslides, and debris-flow modelling. Despite the huge amount of research that has gone into describing the behaviour of granular flows, a constitutive equation that describes the overall behaviour of a flowing granular material is still lacking. The initiation and propagation of submarine granular flows depend mainly on the slope, density, and quantity of the material destabilised. Although certain macroscopic models are able to capture the simple mechanical behaviours, the complex physical mechanisms that occur at the grain scale, such as the hydrodynamic instabilities, the formation of clusters, collapse, and transport, have largely been ignored~\shortcite{Topin2011}. The momentum transfer between the discrete and the continuous phases significantly affects the dynamics of the flow~\shortcite{Peker2007}. Grain-scale description of the granular material enriches the macro-scale variables,  which poorly account for the local rheology of the materials.  In order to describe the mechanism of saturated and/or immersed granular flows, it is important to consider both the dynamics of the solid phase and the role of the ambient fluid~\shortcite{Denlinger2001}. In particular, when the solid phase reaches a high volume fraction, it is important to consider the strong heterogeneity arising from the contact forces between the grains, the drag interactions which counteract the movement of the grains, and the hydrodynamic forces that reduce the weight of the solids inducing a transition from dense compacted to a dense suspended flow~\shortcite{Meruane2010}. The case of the collapse in presence of an interstitial fluid has been less studied. In this paper, we study the effect of permeability and initial density on the run-out evolution.

@@

The Lattice Boltzmann Method is a `micro-particle' based numerical time-stepping procedure for the solution of incompressible fluid flows. Consider a 2D incompressible fluid flow with density $\rho$ and kinematic viscosity \textit{v}, in a rectangular domain \textit{\textbf{D}}. The fluid domain is divided into a rectangular grid or lattice, with the same spacing \textit{`h'} in both the \textit{x-} and the \textit{y-}directions. The present study focuses on two-dimensional problems, hence the \textit{D2Q9} momentum discretisation is adopted (see \shortciteN{He1997} for naming convention). The Multiple Relaxation Time LBM (LBM-MRT) is adopted in the present study for numerical stability. Further details about the modelling can be found in~\shortciteN{Kumar2015,Kumar2017}.

@@

\section{UNDERWATER GRANULAR FLOWS}
In this study, a 2D poly-disperse system ($d_{max}/d_{min} = 1.8$) of circular
discs in fluid was used to understand the behaviour of granular flows on
inclined planes (see~\Cref{fig:setup}). The soil column was modelled using
1000 discs of density \SI{2650}{\kg\per\cubic\meter} and a contact friction
angle of \SI{26}{\degree}. The collapse of the column was simulated inside a
fluid with a density of \SI{1000}{\kg\per\cubic\meter}  and a kinematic
viscosity of \SI{1e-6}{\square\meter\per\second}. The choice of a 2D geometry
has the advantage of cheaper computational effort than a 3D case, making it
feasible to simulate very large systems. A granular column of aspect ratio
`a' of 0.8 was used. Dry analyses were also performed to understand the effect
of hydrodynamic forces on the run-out distance in submerged granular flows.

@@

\shortciteN{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), is compared with a loose
granular column (79\% packing fraction). The permeability is varied by changing
the hydrodynamic radius from 0.7 \textit{R} (high permeability) to 0.95 \textit{R}
(low permeability). A hydrodynamic radius refers to the reduced grain diameter
used only in the LBM computation to simulate a 3D flow through a 2D system of rods.
See~\shortciteN{Kumar2015} for further details on the hydrodynamic radius.

@@

At a high permeability (\textit{r} = 0.7 \textit{R}), the evolution of run-out 
at the initial stage is quicker, which means that the negative pore-pressure 
that is developed during the shearing along the shear-failure surface is 
dissipated faster. Even though the negative pore-pressure is dissipated, due to 
the development of negative pore-pressure the evolution of run-out in fluid is 
slower than its dry counterpart. The rate of pore-pressure dissipation 
decreases with decrease in the permeability. This can be observed by a flatter 
slope in the run-out evolution with decrease in the 
permeability.~\Cref{fig:PWP_ini_dense} shows the distribution of pore-pressure 
in high and low permeability granular media along the horizontal direction at a 
height of $10 \times d$ from the base. In LBM, the pore-pressure in fluid is a 
function of fluid density distribution functions. At time $t = \tau_c$, defined
as the critical time for the flow to be fully mobilised, the 
highly permeable (\textit{r} = 0.7 \textit{R}) granular column 
shows smaller negative pore-pressure in comparison to large negative 
pore-pressures observed in the shearing zone of a low permeable column 
(\textit{r} = 0.9 \textit{R}). This shows that not only does it take longer for the 
pore-pressure to dissipate with a decrease in permeability, but also results in 
almost twice the negative pore-pressure than what is observed in the high 
permeable case (\cref{fig:r095_PWP_ini_dense}). A 
high value of positive pore-pressure is observed at the face of the column 
in the low permeable condition. This indicates that the low permeable column 
fails as a continuous block undergoing a shear failure, which generates a very 
large negative pore-pressure along the failure surface. However, in the case of 
the high permeable column the failure is more localised with multiple negative 
pore-pressure spikes (\cref{fig:r07_PWP_ini_dense}).

@@

\Cref{fig:Packing_Density_a08_dense} shows that the high permeability granular
column shows lower water entrainment (smaller decrease in packing denisty),
which indicates that for highly 
permeable flows the drag force acting on the soil grains predominates over the 
lubrication effect on the run-out behaviour.~\Cref{fig:Fr_a08_dense} shows the 
evolution of Froude's number with time for different permeability conditions. 
In dense granular columns, the Froude's number increases with decrease in 
permeability, however the Froude's number is below a critical value of 0.4 even 
for a low permeable column. Hence, no hydroplaning is observed. In both the low 
and high permeable granular flows, the granular material consolidates at the 
final stage of the flow (\cref{fig:Packing_Density_a08_dense}). This can be 
observed by the increase in the packing density at the final stage due to 
settlement of grains and expulsion of entrained water.

@@

The normalised run-out evolution with time for a loose initial packing (79\% 
packing fraction) with different hydrodynamic radii 0.7 \textit{R} to 0.95
\textit{R} are presented in~\cref{fig:Runout_a08_loose}. The 
run-out evolution a column of grains in suspension is compared with the dry and 
submerged granular columns to understand the influence of hydrodynamic forces 
on the flow kinematics. Similar to the dense granular column, the run-out 
distance increases with increase in the hydrodynamic radius (i.e., decrease in 
permeability). At low permeabilities (\textit{r} = 0.9 and 0.95 \textit{R}), 
the run-out distance is longer than the dry condition. This shows that the 
lubrication effect in low permeability conditions overcomes the influence of 
the drag force and the development of large negative pore-pressure resulting in 
a longer run-out distance. Although the suspended granular masses experience 
higher drag forces and turbulent effects, the run-out evolves almost at the 
same rate in comparison with granular columns with high permeability. This 
shows the effect of permeability on the dissipation rate of negative 
pore-pressure developed during the initial stage of collapse.

@@

Two-dimensional LB-DEM simulations were performed to understand the behaviour of
submarine granular flows. Unlike dry granular collapse, the run-out behaviour in
fluid is dictated by the initial volume fraction. 
\shortciteN{Rondon2011} experimentally observed that the collapse of a granular
column in a viscous fluid is mainly controlled by the initial volume fraction and
not by the aspect ratio of the column. The role of the initial volume fraction 
observed explains the pore pressure feedback mechanism proposed 
by~\shortciteN{Schaeffer2008} in the context of landslides. The dense column
needs to dilate in order to flow. When it starts to fall, liquid is then sucked
into the column, which is then stabilised by the additional viscous drag. By
opposition the loose column when it starts flowing expands and ejects liquid,
leading to a partial fluidisation of the material.
